---
title: "Income, Education, and Race Across Florida Census Tracts"
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(sf)
library(leaflet)
library(htmltools)
library(scales)

# Load preprocessed 2022 5-Year ACS tract data
fl_tracts <- read_rds("data/fl_tracts_acs.rds")

# Prepare hover tooltip labels
labels <- sprintf(
  "<strong>%s</strong><br/>
   Median Household Income: %s<br/>
   Bachelor's Degree or Higher: %s%%<br/>
   <hr style='margin: 4px 0;'/>
   Hispanic / Latino: %s%%<br/>
   White: %s%%<br/>
   Black: %s%%<br/>
   Asian: %s%%",
  fl_tracts$NAME,
  ifelse(is.na(fl_tracts$med_income), "Data unavailable", dollar(fl_tracts$med_income)),
  ifelse(is.na(fl_tracts$pct_bachelor), "N/A", fl_tracts$pct_bachelor),
  ifelse(is.na(fl_tracts$pct_hispanic), "N/A", fl_tracts$pct_hispanic),
  ifelse(is.na(fl_tracts$pct_white), "N/A", fl_tracts$pct_white),
  ifelse(is.na(fl_tracts$pct_black), "N/A", fl_tracts$pct_black),
  ifelse(is.na(fl_tracts$pct_asian), "N/A", fl_tracts$pct_asian)
) |> lapply(HTML)

# Color palettes for each variable
pal_income   <- colorNumeric("YlGnBu", domain = fl_tracts$med_income, na.color = "#d9d9d9")
pal_bachelor <- colorNumeric("Purples", domain = fl_tracts$pct_bachelor, na.color = "#d9d9d9")
pal_hispanic <- colorNumeric("YlOrRd", domain = fl_tracts$pct_hispanic, na.color = "#d9d9d9")
pal_black    <- colorNumeric("PuRd", domain = fl_tracts$pct_black, na.color = "#d9d9d9")
pal_asian    <- colorNumeric("Greens", domain = fl_tracts$pct_asian, na.color = "#d9d9d9")
pal_white    <- colorNumeric("Blues", domain = fl_tracts$pct_white, na.color = "#d9d9d9")
```

## Geographic Patterns Across Florida Census Tracts

The interactive map below allows you to explore how median household income, educational attainment, and racial/ethnic compositions vary across Florida's census tracts.

Use the radio buttons in the top-right corner to toggle layers. Hover over any census tract to inspect its specific figures.

```{r}
#| label: interactive-tract-map
#| message: false
#| warning: false
#| echo: false

leaflet(fl_tracts, width = "100%", height = "560px") |>
  addProviderTiles(providers$Esri.WorldGrayCanvas) |>
  addPolygons(
    group = "Median Household Income",
    fillColor = ~pal_income(med_income),
    fillOpacity = 0.7, weight = 0.4, color = "#666666",
    label = labels,
    highlightOptions = highlightOptions(weight = 2, color = "#111111", bringToFront = TRUE)
  ) |>
  addPolygons(
    group = "% Bachelor's Degree+",
    fillColor = ~pal_bachelor(pct_bachelor),
    fillOpacity = 0.7, weight = 0.4, color = "#666666",
    label = labels,
    highlightOptions = highlightOptions(weight = 2, color = "#111111", bringToFront = TRUE)
  ) |>
  addPolygons(
    group = "% Hispanic / Latino",
    fillColor = ~pal_hispanic(pct_hispanic),
    fillOpacity = 0.7, weight = 0.4, color = "#666666",
    label = labels,
    highlightOptions = highlightOptions(weight = 2, color = "#111111", bringToFront = TRUE)
  ) |>
  addPolygons(
    group = "% Black",
    fillColor = ~pal_black(pct_black),
    fillOpacity = 0.7, weight = 0.4, color = "#666666",
    label = labels,
    highlightOptions = highlightOptions(weight = 2, color = "#111111", bringToFront = TRUE)
  ) |>
  addPolygons(
    group = "% Asian",
    fillColor = ~pal_asian(pct_asian),
    fillOpacity = 0.7, weight = 0.4, color = "#666666",
    label = labels,
    highlightOptions = highlightOptions(weight = 2, color = "#111111", bringToFront = TRUE)
  ) |>
  addPolygons(
    group = "% White",
    fillColor = ~pal_white(pct_white),
    fillOpacity = 0.7, weight = 0.4, color = "#666666",
    label = labels,
    highlightOptions = highlightOptions(weight = 2, color = "#111111", bringToFront = TRUE)
  ) |>
  addLayersControl(
    baseGroups = c(
      "Median Household Income",
      "% Bachelor's Degree+",
      "% Hispanic / Latino",
      "% Black",
      "% Asian",
      "% White"
    ),
    options = layersControlOptions(collapsed = FALSE)
  )
```

---

## Comparing Race/Ethnicity with Household Income

The scatterplots below examine how tract-level racial and ethnic compositions correlate with median household income across Florida census tracts. Each point represents one tract, and the blue line shows the linear trend.

```{r}
#| label: income-scatterplots
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 4.5

# Drop geometry and prepare data for plotting
tract_data <- fl_tracts |>
  st_drop_geometry() |>
  drop_na(med_income, pct_bachelor, pct_hispanic, pct_black, pct_asian)

income_long <- tract_data |>
  select(med_income, pct_hispanic, pct_black, pct_asian) |>
  pivot_longer(
    cols = c(pct_hispanic, pct_black, pct_asian),
    names_to = "group",
    values_to = "pct"
  ) |>
  mutate(
    group = case_match(
      group,
      "pct_hispanic" ~ "% Hispanic / Latino",
      "pct_black"    ~ "% Black",
      "pct_asian"    ~ "% Asian"
    )
  )

ggplot(income_long, aes(x = pct, y = med_income)) +
  geom_point(alpha = 0.2, color = "#2c3e50", size = 1.2) +
  geom_smooth(method = "lm", formula = y ~ x, color = "#2980b9", se = FALSE, linewidth = 1) +
  facet_wrap(~group, scales = "free_x") +
  scale_y_continuous(labels = dollar_format()) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(
    title = "Tract-Level Race/Ethnicity vs. Median Household Income",
    subtitle = "U.S. Census Bureau, 2018–2022 5-Year ACS (Florida)",
    x = "Demographic Group (% of Tract Population)",
    y = "Median Household Income"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold")
  )
```

---

## Comparing Race/Ethnicity with Educational Attainment

The scatterplots below examine the relationship between demographic compositions and educational attainment, measured as the percentage of adults age 25 and older with a bachelor's degree or higher.

```{r}
#| label: education-scatterplots
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 4.5

edu_long <- tract_data |>
  select(pct_bachelor, pct_hispanic, pct_black, pct_asian) |>
  pivot_longer(
    cols = c(pct_hispanic, pct_black, pct_asian),
    names_to = "group",
    values_to = "pct"
  ) |>
  mutate(
    group = case_match(
      group,
      "pct_hispanic" ~ "% Hispanic / Latino",
      "pct_black"    ~ "% Black",
      "pct_asian"    ~ "% Asian"
    )
  )

ggplot(edu_long, aes(x = pct, y = pct_bachelor)) +
  geom_point(alpha = 0.2, color = "#34495e", size = 1.2) +
  geom_smooth(method = "lm", formula = y ~ x, color = "#8e44ad", se = FALSE, linewidth = 1) +
  facet_wrap(~group, scales = "free_x") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(
    title = "Tract-Level Race/Ethnicity vs. Educational Attainment",
    subtitle = "Share of adults age 25+ with a bachelor's degree or higher vs. demographic percentage",
    x = "Demographic Group (% of Tract Population)",
    y = "Adults with Bachelor's Degree or Higher"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold")
  )
```

---

## What Does the Data Show?

An empirical evaluation of Florida's census tracts reveals several distinct geographic patterns and correlations:

1. **Strong Link Between Education and Income**: Across all census tracts, there is a strong positive correlation ($r = 0.70$) between bachelor's degree attainment and median household income. Tracts where a large majority of adults hold degrees consistently report median household incomes well exceeding \$100,000.
2. **Geographic Distribution of Income and Education**: 
   - High-income and high-education tracts heavily cluster in coastal communities (such as coastal Sarasota, coastal Collier, and coastal Palm Beach) as well as planned suburban corridors around Orlando, Tampa, and Jacksonville.
   - Lower median income and lower bachelor's degree attainment are concentrated in inland rural agricultural tracts (such as in Hendry, Hardee, and DeSoto counties) and in select urban cores.
3. **Demographic Group Patterns**:
   - **Asian**: Higher percentages of Asian residents in a tract show positive correlations with both median household income ($r = 0.21$) and educational attainment ($r = 0.29$). High-percentage Asian tracts cluster around major university centers (e.g., Gainesville, East Orlando near UCF, Tallahassee) and suburban tech/medical hubs.
   - **Black**: Tracts with higher percentages of Black residents show negative correlations with median household income ($r = -0.34$) and bachelor's degree attainment ($r = -0.35$). These tracts are geographically concentrated in historically distinct neighborhoods within metropolitan areas (such as Jacksonville, Miami, and Tampa) as well as specific rural agricultural areas in North and Central Florida.
   - **Hispanic / Latino**: Statewide correlations between % Hispanic population and income ($r = -0.08$) and education ($r = -0.11$) are near zero. This reflects significant internal diversity across Florida's Hispanic communities: tracts in areas like Doral and Weston exhibit high median incomes and bachelor's degree rates, while tracts in South Dade and agricultural regions reflect lower median incomes.
4. **Correlation vs. Causation**: These tract-level correlations describe neighborhood-level associations rather than individual-level causes. Differences across tracts reflect a combination of historical settlement patterns, regional economic bases, and housing costs.
