---
title: "Florida Census Demographics (2000–2020)"
---

```{r}
#| label: setup
#| message: false
#| echo: false

library(tidyverse)
library(sf)
library(leaflet)
library(htmltools)

# Load county boundaries and decennial census data
fl_geo  <- read_rds("data/fl_geo.rds")
fl_race <- read_rds("data/fl_race_2000_2020.rds")

# Reshape data to wide format to compute percentage point changes (2000 to 2020)
race_wide <- fl_race |>
  mutate(pct = round(100 * (value / summary_value), 1)) |>
  select(GEOID, variable, year, pct) |>
  pivot_wider(names_from = c(variable, year), values_from = pct) |>
  mutate(
    diff_hisp  = round(hispanic_2020 - hispanic_2000, 1),
    diff_black = round(black_2020 - black_2000, 1),
    diff_white = round(white_2020 - white_2000, 1),
    diff_asian = round(asian_2020 - asian_2000, 1)
  )

# Combine with county geometries and reproject to WGS84
fl_map <- left_join(fl_geo, race_wide, by = "GEOID") |>
  st_transform(4326)

# Create hover labels summarizing all four groups for each county
labels <- sprintf(
  "<strong>%s</strong><br/>
   Hispanic: %s%% &rarr; %s%% (%+0.1f pts)<br/>
   White: %s%% &rarr; %s%% (%+0.1f pts)<br/>
   Black: %s%% &rarr; %s%% (%+0.1f pts)<br/>
   Asian: %s%% &rarr; %s%% (%+0.1f pts)",
  fl_map$NAME,
  fl_map$hispanic_2000, fl_map$hispanic_2020, fl_map$diff_hisp,
  fl_map$white_2000, fl_map$white_2020, fl_map$diff_white,
  fl_map$black_2000, fl_map$black_2020, fl_map$diff_black,
  fl_map$asian_2000, fl_map$asian_2020, fl_map$diff_asian
) |> lapply(HTML)

# Define color scales for each demographic group
pal_hisp  <- colorNumeric("YlOrRd", domain = fl_map$diff_hisp)
pal_white <- colorNumeric("Blues", domain = fl_map$diff_white)
pal_black <- colorNumeric("Purples", domain = fl_map$diff_black)
pal_asian <- colorNumeric("Greens", domain = fl_map$diff_asian)
```

## Two Decades of Demographic Change Across Florida

Between 2000 and 2020, Florida saw significant demographic shifts across all major racial and ethnic groups:

- **Hispanic**: Substantial increases statewide, especially in central and southwest Florida.
- **White (non-Hispanic)**: Decreased as a overall percentage of population in almost all 67 counties.
- **Black**: Steady population shares in northern counties, with growth in select central and southern metros.
- **Asian**: Steady percentage growth across major suburban and university counties.

Use the radio buttons in the upper right to toggle which demographic shift colors the map. Hover over any county to view the full 20-year shift for all four groups.

```{r}
#| label: interactive-map
#| message: false
#| echo: false

leaflet(fl_map, width = "100%", height = "540px") |>
  addProviderTiles(providers$Esri.WorldGrayCanvas) |>
  addPolygons(
    group = "Hispanic Change",
    fillColor = ~pal_hisp(diff_hisp),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addPolygons(
    group = "White Change",
    fillColor = ~pal_white(diff_white),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addPolygons(
    group = "Black Change",
    fillColor = ~pal_black(diff_black),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addPolygons(
    group = "Asian Change",
    fillColor = ~pal_asian(diff_asian),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addLayersControl(
    baseGroups = c("Hispanic Change", "White Change", "Black Change", "Asian Change"),
    options = layersControlOptions(collapsed = FALSE)
  )
```
